Reduced Nucleotide Diversity on a Plant Y Chromosome Following Recent Recombination Suppression

### 2016-08-10 manuscript.tex
ms edits

\documentclass[9pt,twocolumn,twoside]{gsajnl}
% Use the documentclass option 'lineno' to view line numbers
\usepackage{booktabs}
\usepackage{siunitx}
\usepackage{geometry} % just not to bother with the table width
%\usepackage[colorinlistoftodos]{todonotes} % comments in margins
%\setlength{\marginparwidth}{1.25cm}
%\definecolor{flame}{rgb}{0.89, 0.35, 0.13}
%\newcommand{\josh}[1]{\textcolor{flame}{\emph{\scriptsize {#1}}}}
%\newcommand{\citex}{\textcolor{red}{\bf (CITE)\,}}
%\newcommand{\out}[1]{\textcolor{red}{#1}} %outline placeholder text
\newcommand{\X}{\textcolor{red}{\bf X\,}}

@@

\subsection*{Y-chromosome diversity in \textit{R. hastatulus} is very low}
Our analysis of polymorphism in \textit{R. hastatulus} revealed that diversity on the Y chromosome is significantly lower than neutral predictions, with estimates indicating $Y/A=0.02$, a 12.5 fold reduction compared to the neutrally expected $Y/A = 0.25 (P<0.0001)$. This corresponds to a 50-fold reduction compared to the mean diversity estimated on autosomes (\X), and a 40-fold reduction compared to X chromosome diversity (Table 1). By normalizing X- and Y-chromosome diversity estimates by autosomal diversity, these results suggest that the X-Y difference we observe (an \X fold difference) is not due to some process specifically elevating diversity on the X chromosome. In addition, a recent analysis of sex-linked substitution rates in this species found that the number of synonymous mutations in X and Y lineages, estimated by both parsimony and maximum likelihood, were not significantly different \citep{hough2014}. This further suggests that the low diversity on the Y is not likely due to a lower mutation rate.

\begin{table}
\centering
\caption{Estimates of neutral diversity by race on \textit{R. hastatulus} sex chromosomes and autosomes.}
\begin{tabular}{ccccccccc}
\textbf{} & \multicolumn{2}{c}{\textbf{Texas}} & \multicolumn{2}{c}{\textbf{South Carolina}} & \multicolumn{2}{c}{\textbf{Florida}} \\
chromosome & $\theta$ & $\theta/\theta_{A}$ & $\theta$ & $\theta/\theta_{A}$ & $\theta$ & $\theta/\theta_{A}$ \\
\midrule
A & 0.006 & 1 & 0.006 & 1 & 0.005 & 1 \\    
X & 0.0047 & 0.85 & 0.002 & 0.33 & 0.0047 & 0.37 \\ 
Y & $\num{e-4}$ & 0.002 & $\num{e-4}$ & 0.002 & $\num{e-4}$ & 0.002 \\ 
\addlinespace

\bottomrule
\end{tabular}

\end{table}

@@

\subsection*{Male-to-female sex ratio of $N_{e}$}

>>>>

\documentclass[9pt,twocolumn,twoside]{gsajnl}
% Use the documentclass option 'lineno' to view line numbers
\usepackage{booktabs}
\usepackage{siunitx}
\usepackage{adjustbox}
\usepackage{geometry} % just not to bother with the table width
%\usepackage[colorinlistoftodos]{todonotes} % comments in margins
%\setlength{\marginparwidth}{1.25cm}
%\definecolor{flame}{rgb}{0.89, 0.35, 0.13}
%\newcommand{\josh}[1]{\textcolor{flame}{\emph{\scriptsize {#1}}}}
%\newcommand{\citex}{\textcolor{red}{\bf (CITE)\,}}
%\newcommand{\out}[1]{\textcolor{red}{#1}} %outline placeholder text
\newcommand{\X}{\textcolor{red}{\bf X\,}}

@@

\subsection*{Y-chromosome diversity in \textit{R. hastatulus} is very low}
Our analysis revealed that diversity on the \textit{R. hastatulus} Y chromosome is significantly lower than expected under neutrality, with estimates indicating $Y/A=0.02$, which is 12.5 fold lower than the standard neutral prediction of $Y/A = 0.25 (P<0.0001)$, and a 40-fold reduction compared to mean diversity on the X chromosome (Table 1). Note that by normalizing X and Y diversity by autosomal diversity, our results indicate that the X-Y difference we observed was not due to an elevation of X chromosome diversity, but a Y-specific diversity reduction. Conceivably, low diversity on the Y could arise from a low Y-chromosome mutation rate or a lower mutation rate in males compared to females. However, recent evidence indicates that the number of synonymous mutations in X and Y lineages, estimated by both parsimony and maximum likelihood, were not significantly different \citep{hough2014}.

% The pattern of reduced diversity on the Y chromosome is observed in both Africans and Europeans, suggesting that the effect is not population-specific, and holds regardless of whether the neutral sequence analyzed is near or far from genes (Table 1). Previous

%Diversity on the Y chromosome is likely not being under-estimated due to the inability to call variants in haploid regions of the genome because diversity on the X measured in females, where the X is diploid, is nearly identical to diversity on the X measured in males, where the X is haploid (Figure S2).

\begin{table}[t!]
\centering
\caption{Estimates of neutral diversity by race on \textit{R. hastatulus} sex chromosomes and autosomes.}
\begin{tabular}{ccccccccc}
\textbf{} & \multicolumn{2}{c}{\textbf{Texas}} & \multicolumn{2}{c}{\textbf{South Carolina}} & \multicolumn{2}{c}{\textbf{Florida}} \\
chromosome & $\theta$ & $\theta/\theta_{A}$ & $\theta$ & $\theta/\theta_{A}$ & $\theta$ & $\theta/\theta_{A}$ \\
\midrule
A & 0.006 & 1 & 0.006 & 1 & 0.005 & 1 \\    
X & 0.0047 & 0.85 & 0.002 & 0.33 & 0.0047 & 0.37 \\ 
Y & $\num{e-4}$ & 0.002 & $\num{e-4}$ & 0.002 & $\num{e-4}$ & 0.002 \\ 
\addlinespace

\bottomrule
\end{tabular}
\end{table}

%In contrast to the Texas sex chromosome race, our analysis revealed that Y-chromosome diversity was considerably higher in the North Carolina race (see Supplementary Material Figure S1). This difference could have arisen due to population substructure, resulting in an elevated overall Y diversity in the pooled sample. To test this hypothesis, we conducted a phylogenetic analysis of sex-linked sequences from all populations in our study. The resulting sequence phylogeny (Figure 4) revealed strong support for the Y-linked genes from the XY1Y2 race being paraphyletic (98% bootstrap support). In particular, our analysis revealed that samples from two populations, both from South Carolina (hereafter the SC sub-clade), formed a monophyletic group, whereas samples from Florida and Georgia (hereafter the FL sub-clade) were also monophyletic, but were more closely related to the XY Texas race. Although our sampling here is limited, these results suggest that introgression may have occurred between the ancestral Texas (XY race) and the derived North Carolina (XY1Y2) race, leading to a second, derived, XY1Y2 sub-clade and an elevated level of Y-diversity in our pooled sample. As these races are inter-fertile (Smith reference about hybrid zone), we suggest that the SC sub-clade likely originated through hybridization between a female from the FL clade harboring the X-autosome fusion, and a male from the XY race. Notably, when we separated samples by the two inferred sub-clades of Y–linked sequences, the estimated level of Y diversity was consistent across all three clades, suggesting that the extensive reduction in diversity on this chromosome is not a population or race-specific phenomenon (see Supplementary Material, Figure S1).

%Our data indicate a significant reduction in X/A polymorphism in the derived SC and FL sub-clades of the North Carolina race (X/AFL=0.33 and X/ASC=0.37) compared to the Texas race (X/ATX=0.85). This reduction in diversity may be associated with the recent origin of the XY1Y2 sex chromosome system that occurs in these two sub-clades. The XY1Y2 sex chromosome system is thought to have originated through an X-autosome fusion (Smith 1964) involving the ancestral 3rd chromosome in the Texas race. Evidence supporting this autosomal origin was recently obtained by Grabowska-Joachimiak et al. (2014), who reported that the ancestral third chromosome in the Texas race carries the 5S rDNA locus, which is now found on both the neo-X and the Y2 sex chromosomes in the derived North Carolina race. That this fusion is likely to have been recent is suggested by C-banding/DAPI experiments showing that while the ancestral Y chromosome is highly heterochromatic, the translocated neo-Y is still euchromatic (Grabowska-Joachimiak et al. 2014), and also shows less genetic degeneration than the ancestral Y chromosome (Hough et al. 2014). Therefore, if recent positive selection was responsible for driving the evolution of this fusion, then the formerly autosomal segment on the X chromosome in the XY1Y2 sub-clades (the neo-X) may have experienced a selective sweep. This process may explain the reduction in diversity that we observed in the derived XY1Y2 sub-clades, though further work is needed to determine the factors driving the establishment of the X-autosome fusion.

%We next consider several possible models that could explain this unexpectedly low amount of diversity found on the Y chromosome relative to other genomic regions. Such models include differences in the variance in reproductive success between males and females, purifying selection on the Y chromosome, and positive selection on the Y chromosome.

\subsection*{Female biased sex ratios and variance in male fitness}

@@
